\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{physics}
\usepackage{array}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}

\title{Vector Databases \& Embeddings}
\author{Saraansh Wadkar}
\date{}

\begin{document}

\maketitle

\section{What Is a Vector?}

In the context of embeddings and vector databases, a vector is an
$n$-dimensional ordered list of numbers:
\[
\mathbf{v} = [v_1, v_2, \dots, v_n].
\]
In practice, each coordinate is usually stored as a signed 32-bit
floating-point value, so a single vector is often represented as an
$n$-dimensional array of 4-byte floats.

An embedding vector is useful because it encodes semantic meaning into
geometry. Rather than storing meaning in words or symbols directly, the
model maps text, images, or other data into a point in a high-dimensional
space. Items with similar meaning are placed closer together, while items
with different meanings are placed farther apart.

For example, two sentences about databases may produce vectors that point
in similar directions, even if they do not share the exact same words.
This is what makes vector search possible: we compare the geometry of
embeddings instead of relying only on exact keyword matches.

\subsection{How Embeddings Are Created}

The key question is how a piece of text becomes a useful vector at all.
An embedding model first breaks text into smaller units called tokens.
Those tokens are then processed by a neural network that has been trained
on very large amounts of data. During training, the model learns internal
patterns about which words, phrases, and concepts tend to appear in
similar contexts.

By the time the model is used for embeddings, it has learned to map text
into a high-dimensional latent space. This means that the output numbers
are not chosen by hand and do not have obvious human-readable meanings on
their own. Instead, the full vector matters as a whole. The position of
that vector reflects the relationships the model learned during training.

For example, suppose we embed the sentence ``the cat sat on the mat.'' The
model first tokenizes it into pieces such as ``the'', ``cat'', ``sat'',
``on'', ``the'', and ``mat''. Those tokens are passed through many layers
of the network, which combine them into a single output vector such as
\[
\mathbf{e}_{\text{cat}} = [0.12, -0.44, 0.91, \dots, 0.08].
\]
The exact numbers are not human-interpretable on their own. What matters
is that another sentence like ``a kitten is resting on a rug'' will likely
produce a vector in a nearby region of the space, because the model has
learned that ``cat'' and ``kitten'' are related concepts and that both
sentences describe similar situations.

As a result, text with similar meaning tends to produce vectors that end
up near each other, even when the wording changes. The embedding works
because the model has compressed many layers of statistical and semantic
structure into a numerical representation that preserves useful similarity
relationships.

\subsection{Semantic Search Example}

Suppose you are shopping for a car and want something good for
off-road adventures. You have a small catalog of cars, each with a short
description. At first this sounds like a simple search problem, but there
is an immediate issue: none of the descriptions may literally contain the
phrase ``off-road adventures.'' A plain keyword search or \texttt{Ctrl+F}
style lookup would therefore miss relevant cars even if some of them are
actually a good fit.

Here is the dataset:

\begin{enumerate}
    \item ``Tesla Model 3 is an electric sedan designed for daily commuting with great efficiency''
    \item ``Ford F-150 is a powerful pickup truck built for hauling and towing heavy loads''
    \item ``Toyota RAV4 is a compact SUV with good fuel economy and family-friendly features''
    \item ``Porsche 911 is a high-performance sports car built for speed and handling''
    \item ``Honda Civic is a reliable compact car ideal for commuting''
    \item ``Chevy Bolt is an affordable electric hatchback for city driving''
\end{enumerate}

The Ford F-150 and Toyota RAV4 might both be relevant to an
``off-road adventures'' style query, but neither description uses those
exact words. So how do we search by meaning instead of by exact text?

The key idea is to embed each description into a vector. Conceptually,
each record might look like:

\begin{verbatim}
{
  car_name: "Tesla Model 3",
  description: "Electric sedan for efficient daily commuting",
  vector: [...] 
}
\end{verbatim}

The important point is that the \texttt{vector} is produced from the
\texttt{description}. Once all the descriptions are embedded, we do the
same thing to the user's query:

\begin{quote}
``vehicle good for off-road adventures''
\end{quote}

In this situation, the Toyota RAV4 and Ford F-150 might rank higher than
the Tesla Model 3 or Honda Civic, even though none of the descriptions
contain the exact phrase ``off-road adventures.'' The system works because
the embedding model has placed related meanings in nearby parts of the
vector space. This is the core promise of semantic search: retrieve by
meaning, not just by literal keyword overlap.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=2.4, >=Latex]
    \draw[->, thin] (-1.2,0) -- (1.3,0);
    \draw[->, thin] (0,-1.0) -- (0,1.2);
    \node[below right] at (1.28,0) {\scriptsize $x$};
    \node[above left] at (0,1.18) {\scriptsize $y$};

    \draw[->, very thick, red!75!black!35] (0,0) -- (0.98,-0.10);
    \node[red!75!black!55, right] at (0.98,-0.10) {\scriptsize Tesla Model 3};

    \draw[->, very thick, blue!70!black!35] (0,0) -- (0.42,0.91);
    \node[blue!70!black!55, above right] at (0.42,0.91) {\scriptsize Toyota RAV4};

    \draw[->, very thick, green!50!black!35] (0,0) -- (0.10,1.00);
    \node[green!50!black!55, above] at (0.10,1.00) {\scriptsize Ford F-150};

    \draw[->, very thick, orange!90!black!35] (0,0) -- (0.84,-0.54);
    \node[orange!90!black!55, below right] at (0.84,-0.54) {\scriptsize Porsche 911};

    \draw[->, very thick, teal!70!black!35] (0,0) -- (0.82,0.44);
    \node[teal!70!black!55, right] at (0.82,0.44) {\scriptsize Honda Civic};

    \draw[->, very thick, purple!70!black!35] (0,0) -- (0.97,0.16);
    \node[purple!70!black!55, right] at (0.97,0.16) {\scriptsize Chevy Bolt};

    \draw[->, ultra thick, black] (0,0) -- (-0.08,1.00);
    \node[black, left, align=right] at (-0.52,0.92) {\scriptsize query:\\[-1pt]\scriptsize off-road adventures};
\end{tikzpicture}
\caption{A 2D picture of semantic similarity. Real embeddings are
high-dimensional, but the same directional intuition applies.}
\end{figure}

This picture is only 2D, while real embeddings live in much higher
dimensions, but it captures the basic idea. Cars that are close together
have similar descriptions. The Tesla Model 3, Honda Civic, and Chevy Bolt
cluster in a more commuting- and city-driving-oriented direction. The
Toyota RAV4 and Ford F-150 sit much closer to each other because their
descriptions both emphasize utility. The query vector for
``off-road adventures'' also points in that same general direction, which
is why those vehicles would be better matches.

At that point, the natural question becomes: how can we mathematically
find which vectors are most similar?

\subsubsection{Cosine Similarity}

Once we have a query vector and a set of car vectors, we need a
mathematical way to ask: which vectors point in the most similar
direction? One of the most common answers is cosine similarity. Given
vectors $\mathbf{u}$ and $\mathbf{v}$, their cosine similarity is
\[
\cos(\theta) =
\frac{\mathbf{u} \cdot \mathbf{v}}{\norm{\mathbf{u}} \norm{\mathbf{v}}}.
\]

Intuitively, this is about angle. If two vectors point in almost the same
direction, then the angle between them is close to $0^\circ$, and
$\cos(0^\circ) = 1$. That means the vectors are highly similar. If the
angle is around $90^\circ$, then $\cos(90^\circ) = 0$, which means they
are effectively unrelated. If they point in opposite directions, the angle
approaches $180^\circ$ and the cosine becomes negative.

We can make that concrete by isolating the query vector and comparing it
against a few individual cars:

\begin{figure}[h]
\centering
\begin{minipage}[t]{0.31\textwidth}
\centering
\begin{tikzpicture}[scale=1.4, >=Latex]
    \draw[->, thin] (-0.1,0) -- (1.1,0);
    \draw[->, thin] (0,-0.1) -- (0,1.1);
    \node[below right] at (1.08,0) {\scriptsize $x$};
    \node[above left] at (0,1.08) {\scriptsize $y$};
    \draw[->, ultra thick, black] (0,0) -- (-0.06,1.0);
    \draw[->, very thick, green!50!black] (0,0) -- (0.17,0.98);
    \node[gray!70!black] at (0.34,0.83) {\scriptsize $\Delta \theta \approx 10^\circ$};
    \node[black, above] at (0,1.0) {\scriptsize query};
    \node[green!50!black, above right] at (0.17,0.98) {\scriptsize Ford F-150};
\end{tikzpicture}

\vspace{0.2em}
\scriptsize High similarity: $\cos(10^\circ) \approx 0.98$. The query and
Ford point in almost the same direction, so this is a strong semantic
match.
\end{minipage}
\hfill
\begin{minipage}[t]{0.31\textwidth}
\centering
\begin{tikzpicture}[scale=1.4, >=Latex]
    \draw[->, thin] (-0.1,0) -- (1.1,0);
    \draw[->, thin] (0,-0.1) -- (0,1.1);
    \node[below right] at (1.08,0) {\scriptsize $x$};
    \node[above left] at (0,1.08) {\scriptsize $y$};
    \draw[->, ultra thick, black] (0,0) -- (-0.06,1.0);
    \draw[->, very thick, purple!70!black] (0,0) -- (0.99,0.14);
    \node[gray!70!black] at (0.47,0.54) {\scriptsize $\Delta \theta \approx 80^\circ$};
    \node[black, above] at (0,1.0) {\scriptsize query};
    \node[purple!70!black, right] at (0.99,0.14) {\scriptsize Chevy Bolt};
\end{tikzpicture}

\vspace{0.2em}
\scriptsize Weak similarity: $\cos(80^\circ) \approx 0.17$. Both are
vehicles, but the meanings are mostly different, so this is close to
noise.
\end{minipage}
\hfill
\begin{minipage}[t]{0.31\textwidth}
\centering
\begin{tikzpicture}[scale=1.4, >=Latex]
    \draw[->, thin] (-0.1,0) -- (1.1,0);
    \draw[->, thin] (0,-0.75) -- (0,1.1);
    \node[below right] at (1.08,0) {\scriptsize $x$};
    \node[above left] at (0,1.08) {\scriptsize $y$};
    \draw[->, ultra thick, black] (0,0) -- (-0.06,1.0);
    \draw[->, very thick, orange!90!black] (0,0) -- (0.80,-0.60);
    \node[gray!70!black] at (0.54,0.30) {\scriptsize $\Delta \theta \approx 125^\circ$};
    \node[black, above] at (0,1.0) {\scriptsize query};
    \node[orange!90!black, below right] at (0.80,-0.60) {\scriptsize Porsche 911};
\end{tikzpicture}

\vspace{0.2em}
\scriptsize Negative similarity: $\cos(125^\circ) \approx -0.57$. This
vector points in a very different direction, so it is a poor match for the
query.
\end{minipage}
\caption{Cosine similarity compares the angle between the query vector and
each car vector. Small angles mean high similarity; large angles mean weak
or even negative similarity.}
\end{figure}

In the car-shopping example, we would compute the cosine similarity
between the query vector and each car description vector, then sort the
results from highest to lowest score. If the query vector points in a
similar direction to the Toyota RAV4 vector, that suggests their meanings
are related even though the query says ``off-road adventures'' and the
description says ``compact SUV.'' This is exactly why cosine similarity is
useful: it lets us retrieve semantically related results without needing
the same literal words to appear in both strings.

In practice, cosine similarity scores are often interpreted as rough
relevance signals. A score around $0.9$ usually indicates a very strong
match. Around $0.7$ is often still meaningfully related, but less precise.
Around $0.4$ is more borderline: there may be some overlap in topic or
context, but the match is weak. Around $0.2$ is often close to noise.
These are heuristics rather than hard rules, since the exact thresholds
depend on the embedding model and the dataset.

The same idea is not limited to language. Vector representations can also
capture similarity for numbers and other structured features. If two data
points have similar numerical patterns, their vectors may still end up
close together, allowing the database to retrieve values that are
numerically or structurally related rather than only textually related.

\end{document}
